---
title: "Creating a DSSAT Weather File"
author: "Lyndon Estes"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DSSAT-weather-files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, collapse = T, 
                      comment = "#>")
```

We will use the `weathdat` dataset, which comes with the `rcropmod`, to provide an example of how to make a DSSAT WTH file. For this we use the `rcropmod` function `weather`. 

## Create weather file 

DSSAT weather files can contain a number of different variables, but at a minimum need daily TMin, TMax, rainfall, and average shortwave radiation. This will allow DSSAT to calculate evapotranspiration using the Priestly-Taylor equation. It can use the more physically realistic FAO56 routine if you can also provide dew point temperature and windspeed. We don't usually have dew point, but the `weather` function draws on `dew_point` to calculate this variable from either relative humidity and temperature or specific humidity and pressure. Here we will use the latter two variables.

```{r, eval = FALSE}
library(rcropmod)
name <- "TEST"
outdir <- "~/DSSAT45/Weather/"
wdat <- weathdat$dat
weathdat$dat[, weather(xy = weathdat$xyz[c("lat", "lon")], 
                       elev = weathdat$xyz["elev"], srad = sw, tmax = tmax,
                       tmin = tmin, prec = prec, wind = wind, sh = sh, 
                       pres = pres, sdate = weathdat$dates["start"], 
                       edate = weathdat$dates["end"], name = name, 
                       outdir = outdir)]
```

## Quality check the WTH file

It's not a bad idea to look at the resulting file to make sure the values make sense, there are missing dates, etc. Here are a few examples of checks with R. 
```{r, eval = FALSE}
wth <- read.table(file = paste0(outdir, "/TEST7932.WTH"), skip = 4, 
                     sep = "", header = TRUE, stringsAsFactors = FALSE)

ymd <- seq(as.Date(weathdat$dates["start"],"%Y%m%d"), 
           as.Date(weathdat$dates["end"],"%Y%m%d"), by = 1)
length(ymd) == length(weathdat$dat[, wind])  # true
nrow(wth) == length(ymd)

# Make sure there are no missing dates - this causes failures
max(sapply(2:length(wth$X.DATE), function(i) {
   as.Date(sprintf("%05s", wth$X.DATE[i]), "%y%j") - 
      as.Date(sprintf("%05s", wth$X.DATE[i - 1]), "%y%j") 
}))
```

And that's it. This file will be used as input in the [DSSAT demonstration tutorial](dssat-simple-demo.html).

